---
title: "Groundwater level state report"
output:
  html_notebook:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: false
    theme: lumen
    fig.cap: TRUE
    df_print: paged
---

```{=html}
<style>
/* Adjust settings for the Table of Contents (TOC) */
#TOC {
    margin: 80px 0 20px -180px; /* Slightly reduced margin for better positioning */
}
div.tocify {
    width: 100%; /* Ensures it adapts well to different screen sizes */
    max-width: 500px;
    max-height: 100vh; /* Ensures TOC never exceeds the viewport height */
    overflow-y: auto; /* Adds scroll if content exceeds max-height */
}

/* Style for the title */
h1.title {
    color: #66b2b2;
    font-weight: bold;
    font-size: 42px;
    text-align: center; /* Centers the title */
    margin-bottom: 20px; /* Adds spacing below the title */
}

/* Body styles */
body {
    text-align: justify;
    font-size: 16pt;
    line-height: 1.6; /* Improves readability with better line spacing */
}

/* Main container styles */
div.main-container {
    max-width: 1280px; /* Reduced slightly for better content width balance */
    margin: 0 auto; /* Centers the container */
    padding: 20px; /* Adds internal spacing */
    box-sizing: border-box; /* Ensures padding doesn’t affect max-width */
}

/* Responsive adjustments */
@media (max-width: 768px) {
    #TOC {
        margin: 50px 0 10px -100px; /* Adjusts TOC margin for smaller screens */
    }
    div.tocify {
        width: 100%;
        max-width: 100%;
    }
    h1.title {
        font-size: 36px; /* Reduces title size for smaller screens */
    }
    body {
        font-size: 14pt; /* Slightly reduces text size for better mobile viewing */
    }
    div.main-container {
        max-width: 95%; /* Allows more flexibility on smaller screens */
        padding: 10px;
    }
}
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(dplyr)
library(RODBC)
library(sf)
library(ggplot2)
library(forcats)
library(tidyr)
library(lubridate)
```

| 
| 

# Load data

## Load data file

This section loads the **main groundwater monitoring dataset** from a structured RData file. The script ensures proper **data formatting** and **metadata assignment** before further analysis.

**Key Dataframes:**

-   **`data_set_clean`**: The primary dataset containing **groundwater level observations**.

-   **`files`**: A list of available groundwater datasets, filtered based on file naming conventions.

```{r , include=FALSE, cache=FALSE}
# Set the target directory

# Set the directory where the file is saved
# As of 16/09/2024 this works
target_dir <- file.path(rprojroot::find_rstudio_root_file(), "2_Data_error_checking_and_cleaning")

state_dir <- file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis")

# Define the pattern to match the files with the desired prefix
pattern <- "^\\d{4}-\\d{2}-\\d{2}_data_set_clean\\.RData$"

# List files in the directory that match the pattern
files <- list.files(path = target_dir, pattern = pattern, full.names = TRUE)

# Check if there are matching files
if (length(files) > 0) {
  # Load the most recent file (if there are multiple, pick the first one)
  load(files[1])
  message("Loaded file: ", files[1])
} else {
  stop("No matching files found.")
}

```

## Load GIS data

This section retrieves and processes **geospatial data** related to groundwater monitoring wells. It reads **shapefiles** and transforms coordinate reference systems (CRS) for compatibility.

**Key Dataframes:**

-   **`allbasins`**: Shapefile of all groundwater basins.

-   **`Regional_boundary_sf`**: Defines regional boundaries.

-   **`Quaternary_aquifers`** and **`Pre_quaternary_aquifers`**: Define aquifer extents.

-   **`SoE_FMUs`**: State of Environment (SoE) Functional Management Units.

-   **`EMA_sf`**: Defines **Environmental Management Areas (EMAs)**.

```{r get_GIS_data}

# Define whether to update data. Set to TRUE if you want to force retrieval from the database.
update_wellstor_data <- FALSE

# Define the GIS layers directory
gis_dir <- file.path(rprojroot::find_rstudio_root_file(), "GIS_layers")

# Find any existing WellStor shapefiles with a date stamp (pattern: "WellStor_YYYYMMDD.shp")
shp_files <- list.files(gis_dir, pattern = "^WellStor_\\d{8}\\.shp$", full.names = TRUE)

if (!update_wellstor_data && length(shp_files) > 0) {
  # If not updating and there are existing shapefiles, pick the most recent one
  file_dates <- sapply(shp_files, function(f) {
    base <- basename(f)
    date_str <- sub("^WellStor_(\\d{8})\\.shp$", "\\1", base)
    as.Date(date_str, format = "%Y%m%d")
  })
  most_recent_file <- shp_files[which.max(file_dates)]
  # The shapefile is assumed to be in NZTM
  WellStor_sf_NZTM <- st_read(most_recent_file)
  message("Loaded WellStor from shapefile: ", most_recent_file)
} else {
  # Either update_wellstor_data is TRUE or no file exists, so retrieve new data
  Con <- RODBC::odbcConnect("HBRCData64", uid = "", pwd = "")
  on.exit(RODBC::odbcClose(Con))
  
  WellStor <- RODBC::sqlQuery(Con, "
    SELECT WELLBore.BoreNo AS well, 
           WELLBore.Easting AS x, 
           WELLBore.Northing AS y, 
           WELLBore.SOE_WL, 
           WELLBore.SOE_WQ
    FROM HbrcDB.dbo.WELLBore WELLBore", 
    stringsAsFactors = FALSE) %>%
    mutate(well = as.factor(well))
  
  # Convert the data frame to an sf object using NZMG coordinates (EPSG:27200)
  WellStor_sf_NZMG <- st_as_sf(WellStor, coords = c("x", "y"), crs = 27200)
  
  # Transform the data to NZTM (EPSG:2193)
  WellStor_sf_NZTM <- st_transform(WellStor_sf_NZMG, crs = 2193)
  
  # Create a date stamp for today's date and define the shapefile path
  date_stamp <- format(Sys.Date(), "%Y%m%d")
  wellstor_shp_path <- file.path(gis_dir, paste0("WellStor_", date_stamp, ".shp"))
  
  # Save the NZTM sf object as a shapefile with the date stamp in the file name
  st_write(WellStor_sf_NZTM, wellstor_shp_path)
  message("WellStor shapefile created and saved: ", wellstor_shp_path)
}

# Ensure that WellStor_sf_NZMG is available by generating it from the NZTM data if needed
if (!exists("WellStor_sf_NZMG")) {
  WellStor_sf_NZMG <- st_transform(WellStor_sf_NZTM, crs = 27200)
  message("Generated WellStor_sf_NZMG from NZTM data")
}

# Ensure that the WellStor attribute dataframe is available
if (!exists("WellStor")) {
  WellStor <- st_drop_geometry(WellStor_sf_NZTM)
  message("Generated WellStor dataframe from NZTM data")
}

# Continue with the rest of your GIS layers and processing...
allbasins <- st_read(file.path(gis_dir, "All_basins7.shp"), crs = 2193) %>% 
  dplyr::select(NAME)

Regional_boundary_sf <- st_read(file.path(gis_dir, "Regional_boundary.shp"))
st_crs(Regional_boundary_sf) <- 2193  # for NZTM
Regional_boundary_sf <- st_transform(Regional_boundary_sf, 2193) %>% 
  st_cast("MULTILINESTRING")

Quaternary_aquifers <- st_read(file.path(gis_dir, "Quaternary_deposits.shp"), crs = 2193)
Pre_qauaternary_aquifers <- st_read(file.path(gis_dir, "Hard_rock_aquifers.shp"), crs = 2193)

bb_box <- st_read(file.path(gis_dir, "bb_box.shp"), crs = 2193) %>% 
  st_cast("MULTILINESTRING")

SoE_FMUs <- st_read(file.path(gis_dir, "SoE_Areas.shp"), crs = 2193) %>% 
  dplyr::select(FMU)

HP_plains_sf <- st_read(file.path(gis_dir, "HP_boundary.shp"))
st_crs(HP_plains_sf) <- 2193  # EPSG for NZTM 2000

RP_plains_sf <- st_read(file.path(gis_dir, "RuaPlains_boundary.shp"))
st_crs(RP_plains_sf) <- 2193  # EPSG for NZTM 2000

EMA_sf <- st_read(file.path(gis_dir, "EMA.shp"))

```

## Load functions

-   The function **`watyr`** assigns each date to the **New Zealand hydrological year**, which runs from **July 1st to June 30th**. This allows for consistent year-to-year comparisons of hydrological conditions.

-   The function **`seayr`** follows a similar structure but shifts the start of the year **one month earlier**, running from **June 1st to May 31st**. This adjustment ensures that full seasons remain within the same seasonal year for more accurate seasonal analysis.

```{r , include=FALSE, cache=FALSE}

#create a function for defining the water year
wtr_yr <- function(dates, start_month=7) {
  # Convert dates into POSIXlt
  dates.posix = as.POSIXlt(dates)
  # Year offset
  offset = ifelse(dates.posix$mon >= start_month - 1, 1, 0)
  # Water year
  adj.year = dates.posix$year + 1899 + offset
  # Return the water year
  adj.year
}


sea_yr <- function(dates, start_month=6) {
  # Convert dates into POSIXlt
  dates.posix = as.POSIXlt(dates)
  # Year offset
  offset = ifelse(dates.posix$mon >= start_month - 1, 1, 0)
  # Water year
  adj.year = dates.posix$year + 1899 + offset
  # Return the water year
  adj.year
}
```

# Prepare data sets

Here, groundwater data is processed for **monthly, seasonal, and percentile-based** analyses. Various transformations and filtering steps are applied to create datasets that enable deeper trend insights.

**Key Dataframes:**

-   **`data_monthly_con`**: Monthly conditions for each well.

-   **`data_seasonal_con`**: Seasonal conditions categorized by hydrological year.

-   **`data_percentile_con`**: Groundwater level percentile classifications.

## Prepare data sets for monthly analysis

-   This section calculates **monthly groundwater statistics** such as:

    -   **Monitoring duration** (`mon_dur`).

    -   **Water year (`watyr`) and seasonal year (`seayr`)** assignments.

    -   Classification into **"Highest-ever", "Lowest-ever", "Normal", "Above-normal", "Below-normal"** categories.

    **Key Dataframes:**

    -   **`data_monthly_con`**: Derived from `data_set_clean`, grouped by **well** and **month**.

```{r , include=FALSE, cache=FALSE}

######################### Monthly groundwater level conditions #############################
data_monthly_con <- data_set_clean %>% 
  dplyr::group_by(well) %>% 
  dplyr::mutate(
    max_date = max(my_month, na.rm = TRUE),
    min_date = min(my_month, na.rm = TRUE),
    cur_date = Sys.Date(),
    mon_dur = round(difftime(max_date, min_date, units = "weeks") / 52.25),
    watyr = wtr_yr(my_month),
    seayr = sea_yr(my_month),
    ) %>% 
  dplyr::ungroup() %>% 
  dplyr::group_by(well, month) %>% 
  dplyr::mutate(
    count = dplyr::n(),
    rank = rank(av_gl, na.last = TRUE),
    con = dplyr::case_when(
      av_gl == max(av_gl, na.rm = TRUE) ~ "Highest-ever",
      av_gl == min(av_gl, na.rm = TRUE) ~ "Lowest-ever",
      av_gl > quantile(av_gl, 0.25, na.rm = TRUE, type = 5) & 
      av_gl < quantile(av_gl, 0.75, na.rm = TRUE, type = 5) ~ "Normal",
      av_gl >= quantile(av_gl, 0.75, na.rm = TRUE, type = 5) & 
      av_gl < max(av_gl, na.rm = TRUE) ~ "Above-normal",
      av_gl <= quantile(av_gl, 0.25, na.rm = TRUE) & 
      av_gl > min(av_gl, na.rm = TRUE) ~ "Below-normal",
      TRUE ~ "NA"
    ),
    month = factor(month, levels = c("July", "August", "September", "October", 
                                     "November", "December", "January", "February", 
                                     "March", "April", "May", "June"))
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::filter(count > 9)  # Must have 10 years or more of data

```

## Prepare data sets for seasonal analysis

-   **`data_season_con`** calculates **seasonal groundwater statistics** by grouping groundwater levels (**GLs**) for each site based on **season** (**Winter, Spring, Summer, Autumn**) and computing the seasonal average.

-   These seasonal averages are then compared to historical seasonal averages and classified into categories such as **Normal, Above-Normal, Below-Normal, Highest-Ever, and Lowest-Ever**.

-   **Key Dataframes:**

    -   **`data_seasonal_con`**: Contains groundwater level summaries per **season**.

```{r, include=FALSE, cache=FALSE}

############################### Seasonal groundwater level conditions ######################
data_season_con <- data_set_clean %>% 
  dplyr::group_by(well) %>% 
  dplyr::mutate(
    watyr = wtr_yr(my_month),
    seayr = sea_yr(my_month),
     ) %>% 
  na.omit() %>% 
  dplyr::group_by(well, seayr) %>% 
  dplyr::filter(dplyr::n_distinct(month) == 12) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    season = forcats::fct_collapse(
      month,
      Spring = c("September", "October", "November"),
      Summer = c("December", "January", "February"),
      Autumn = c("March", "April", "May"),
      Winter = c("June", "July", "August")
    ),
    season = forcats::fct_relevel(season, "Winter", "Spring", "Summer", "Autumn")
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::group_by(well, season, seayr) %>% 
  dplyr::mutate(sea_gl = mean(av_gl, na.rm = TRUE)) %>% 
  dplyr::distinct(well, season, watyr, sea_gl, .keep_all = TRUE) %>% 
  dplyr::ungroup() %>% 
  dplyr::group_by(well, season) %>% 
  dplyr::mutate(
    count = dplyr::n(),
    con = dplyr::case_when(
      sea_gl == max(sea_gl, na.rm = TRUE) ~ "Highest-ever",
      sea_gl == min(sea_gl, na.rm = TRUE) ~ "Lowest-ever",
      sea_gl > quantile(sea_gl, 0.25, na.rm = TRUE) & 
      sea_gl < quantile(sea_gl, 0.75, na.rm = TRUE) ~ "Normal",
      sea_gl >= quantile(sea_gl, 0.75, na.rm = TRUE) & 
      sea_gl < max(sea_gl, na.rm = TRUE) ~ "Above-normal",
      sea_gl <= quantile(sea_gl, 0.25, na.rm = TRUE) & 
      sea_gl > min(sea_gl, na.rm = TRUE) ~ "Below-normal",
      TRUE ~ NA_character_
    )
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(well, sea_gl, seayr, season, con) %>% 
  dplyr::distinct() %>% 
  dplyr::group_by(well) %>% 
  dplyr::mutate(season_count = dplyr::n()) %>% 
  dplyr::filter(season_count > 39) %>%  # Must have more than 10 years of seasonal data (4 seasons per year * 10 years)
  dplyr::ungroup()

#duplicates <- data_season_con[duplicated(data_season_con), ]

```

## Prepare data sets for percentile analysis 1

-   Uses data_set_clean

-   This R code **prepares a dataset (`state`) for analysis** by cleaning and structuring `data_set_clean`,

-   Creates the following dataframes:

    -   state - creates labels for subsequent analysis and restructures some columns

    -   All_GL - uses the state df. Creates a df with 80% data coverage over the monitoring period and minimum of 5 years (5 data points)

    -   WellStor_SF - coverts the WellStor dataframe into an sf object for spatial analysis

    -   WellStor_basins - intersects wellStor and the allbasins shapefile

    -   WellStor_EMA - intersects wellStor and the EMA shapefile

    -   GL_per - merges groundwater level (All_GL) data with WellStor_basins

    -   GL_per_EMA - merges groundwater level (All_GL) data with WellStor_EMA

    -   GL_per_1_EMA - merges groundwater level (data_monthly_con) with WellStor EMA (I need to check what the difference between (All_GL and data_monthly_con is)

    -   GL_per_combined - takes the df GL_per and adds the hydro cateogory "Full record"

```{r , include=FALSE, cache=FALSE}

#Prepare data set for analysis
state <- data_set_clean %>% 
  dplyr::mutate(
    label = paste(well, month, year),
    label1 = paste(well, month),
    month = factor(month, levels = c("January", "February", "March", "April", "May", "June", 
                                     "July", "August", "September", "October", "November", "December")),
    value = GL,  
    year = as.integer(year)) %>% 
  dplyr::distinct() %>%  # Remove exact duplicates
  na.omit()  # Remove rows with NA values

state$label[duplicated(state$label)] #check for duplicate measurements

#Create a dataset with 80% data coverage over the monitoring period and minimum of 5 years (5 data points)
All_GL <- state %>%
  group_by(well) %>%
  mutate(
    end_date = max(datetime), 
    start_date = min(datetime)    ,
    actual_monitoring = round((n() - 1) / 12, 2),
    years_monitored = round(as.numeric((end_date - start_date) / 365.25), 1),
    coverage = actual_monitoring / years_monitored,
    year = as.numeric(year)
  ) %>%
  filter(coverage >= 0.8, years_monitored >= 5) %>%
  mutate(period = "Full") 

#################################

# Convert well locations to spatial data and transform to NZTM
# WellStor_SF <- WellStor_sf %>% 
  # dplyr::select(well)

# Perform spatial intersections
WellStor_basins <- WellStor_sf_NZTM %>% 
  sf::st_intersection(x=allbasins) %>% 
  sf::st_drop_geometry() %>% 
  select(well,NAME)

WellStor_EMA <- WellStor_sf_NZTM %>% 
  sf::st_intersection(x=EMA_sf) %>% 
  sf::st_drop_geometry() %>% 
  select(well, EMA)

# Merge groundwater level (All_GL) data with basin and EMA information
GL_per <- All_GL %>% 
  select(well, datetime, GL) %>% 
  left_join(WellStor_basins, by = "well") %>% 
  mutate(datetime = as.Date(datetime),
         hydro_year = ifelse(format(datetime, "%m") >= 7, 
                             as.numeric(format(datetime, "%Y")), 
                             as.numeric(format(datetime, "%Y")) - 1) %>% as.character()) %>% 
  group_by(well) %>%
  mutate(percentile = percent_rank(GL) * 100) %>%
  ungroup()

# Merge groundwater level (All_GL) data with EMA information

GL_per_EMA <- All_GL %>% 
  select(well, datetime, GL) %>% 
  left_join(WellStor_EMA, by = "well")

GL_per_1_EMA <- data_monthly_con %>% 
  select(well, datetime, GL, watyr) %>% 
  left_join(WellStor_EMA, by = "well")

#########################################################

GL_per_combined <- GL_per %>%
  filter(hydro_year >= "2004") %>%
  bind_rows(mutate(., hydro_year = "Full Record")) %>%
  mutate(hydro_year = factor(hydro_year, levels = c(sort(unique(as.numeric(hydro_year))), "Full Record")))

# Identify the last three hydrological years dynamically
highlight_years <- c(as.character(tail(sort(unique(as.numeric(GL_per_combined$hydro_year))), 3)), "Full Record")
#last_years <- sort(unique(as.numeric(GL_per_filtered$hydro_year)), decreasing = TRUE)[1:3]


```

# Analysis

## Seasonal groundwater conditions by aquifer

-   **Uses the `data_season_con` dataframe**

-   This analysis evaluates **seasonal groundwater level conditions** across **Hawke's Bay from 2019 to 2024**. It does so by **counting how often wells in each aquifer** experienced specific groundwater conditions (e.g., Normal, Above-Normal, Below-Normal) and then **calculates the percentage of occurrences** for each category.

```{r , include=FALSE, cache=FALSE}

################## by Seasonal Year ######################################

# Group seasonal conditions by season/watyr and graph for each water year for each Aquifer area
data_season_2019_2023_by_aquifer <- data_season_con %>%
  mutate(con = fct_relevel(con, "Highest-ever", "Above-normal", "Normal", "Below-normal", "Lowest-ever")) %>%
  ungroup() %>%
  filter(seayr %in% c(2019, 2020, 2021, 2022, 2023)) %>%
  left_join(WellStor_sf_NZTM, by = "well") %>%
  st_as_sf(crs = 2193) %>%  # Ensure the joined data is an sf object with NZTM CRS
  st_intersection(allbasins) %>%
  st_drop_geometry() %>%
  select(well, season, con, NAME, seayr) %>%
  group_by(season, NAME, seayr, con) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(per = round(100 * n / sum(n))) %>%
  na.omit()

```

# Plotting

## Percentile box plot

Box plots visualize groundwater level **distributions across multiple hydrological years**. Shading represents **percentile ranges**, helping detect extreme variations.

**Key Dataframes:**

-   **`boxplot_data`**: Contains percentile-based groundwater statistics.

```{r , include=FALSE, cache=FALSE}
# Create a numeric version of the hydrological year positions (1, 2, 3, ...)
# This numeric value will be used for the x-axis so that the shading can work properly.
GL_per_combined$hydro_year_num <- as.numeric(GL_per_combined$hydro_year)

# Create shading data frame using the numeric positions
shading_df_1 <- data.frame(
  hydro_year_label = levels(GL_per_combined$hydro_year),
  x = 1:length(levels(GL_per_combined$hydro_year))
)
shading_df_1$shade <- rep(c(TRUE, FALSE), length.out = nrow(shading_df_1))
shading_df_1 <- shading_df_1[shading_df_1$shade, ]
shading_df_1$xmin <- shading_df_1$x - 0.5
shading_df_1$xmax <- shading_df_1$x + 0.5

# Now plot using the numeric x position but re-label the x-axis accordingly.
Boxplot <- ggplot(GL_per_combined, aes(x = hydro_year_num, y = percentile, fill = NAME)) +
  # Update the shading rectangle to extend only from y = 0 to 100 (or use -Inf if you prefer)
  geom_rect(data = shading_df_1,
            aes(xmin = xmin, xmax = xmax),
            ymin = 0, ymax = 100,   # Changed ymax to 100 (and optionally ymin to 0)
            fill = "grey70", alpha = 0.5, inherit.aes = FALSE) +
  # Add boxplots with explicit grouping for each hydrological year and NAME
  geom_boxplot(aes(group = interaction(hydro_year, NAME)),
               color = "black", outlier.shape = NA) +
  scale_fill_brewer(palette = "Paired") +
  labs(x = "Hydrological Year", y = "Groundwater Level Percentile (%)") +
  scale_x_continuous(breaks = 1:length(levels(GL_per_combined$hydro_year)),
                     labels = levels(GL_per_combined$hydro_year)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "bottom") +
  guides(fill = guide_legend(title = NULL))


#creates a Map folder
dir.create(file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Box_plots"), recursive = TRUE, showWarnings = FALSE)

# Save the plot inside "Box_plots"
ggsave(
  filename = file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Box_plots", "Percentile_boxplot.png"),
  plot = Boxplot,
  width = 10,
  height = 6,
  dpi = 300
)
```

## Envelope plots - option 1

Envelope plots illustrate **groundwater variability over time** by displaying **percentile-based bands**.

**Key Dataframes:**

-   **`envelope_plot_data`**: Summarizes **groundwater levels for ribbon plot visualizations**.

```{r, include=FALSE, cache=FALSE}

#################### option 2 ##########################################

# Convert hydro_year to numeric before summarisation
GL_per_combined_1 <- GL_per_combined %>%
  mutate(hydro_year = as.numeric(as.character(hydro_year)))

# Summarise data to get percentiles for ribbon plot
summary_data <- GL_per_combined_1 %>%
  filter(NAME == "Heretaunga Plains") %>%
  group_by(hydro_year) %>%
  summarise(
    p10 = quantile(percentile, 0.10, na.rm = TRUE),
    p25 = quantile(percentile, 0.25, na.rm = TRUE),
    p50 = quantile(percentile, 0.50, na.rm = TRUE),  # Median
    p75 = quantile(percentile, 0.75, na.rm = TRUE),
    p90 = quantile(percentile, 0.90, na.rm = TRUE)
  )

# Option 2 - Create the ribbon plot
Envelope <- ggplot(summary_data, aes(x = hydro_year + 1)) +  # Shift x-axis forward by 1 year
  geom_ribbon(aes(ymin = p25, ymax = p75, fill = "25th-75th Percentile"), alpha = 0.5) +  
  geom_ribbon(aes(ymin = p10, ymax = p90, fill = "10th-90th Percentile"), alpha = 0.3) +  
  geom_line(aes(y = p50, colour = "Median"), linewidth = 1.2) +  
  scale_fill_manual(values = c("25th-75th Percentile" = "#66c2a5", "10th-90th Percentile" = "#8da0cb")) +
  scale_colour_manual(values = c("Median" = "black")) +
  labs(
    x = "Hydrological Year", 
    y = "Groundwater Level Percentile (%)",
    fill = "Percentile Ranges",
    colour = NULL
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "bottom"
  ) +
  scale_x_continuous(
    breaks = summary_data$hydro_year + 1,  # Adjust breaks for new x-axis
    labels = function(x) paste0(x - 1, "-", x)  # Correct "YYYY-YYYY" formatting
  )

#creates a Map folder
dir.create(file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Envelope_plot1"), recursive = TRUE, showWarnings = FALSE)


# Save the plot inside "Envelope_plot1"
ggsave(
  filename = file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Envelope_plot1", "Percentile_envelope.png"),
  plot = Envelope,
  width = 10,
  height = 6,
  dpi = 300
)


```

## Envelope plots - options 2

-   Uses data_set_clean

-   This R script **generates and saves groundwater level plots** for **each well** in the dataset. It stores them in a new folder called Envelope_plots

-   The full time series is plotted but the Envelope_data df could be filtered to look at the last X years if desired

```{r , include=FALSE, cache=FALSE}

#Obtain data for plotting
Envelope_data <- data_set_clean %>% 
  group_by(well, month) %>% 
  mutate(minGL= min(GL, na.rm = TRUE),
         maxGL = max(GL, na.rm = TRUE),
         Q20 = quantile(GL,0.2, na.rm=TRUE),
         Q80 = quantile(GL,0.8, na.rm=TRUE))

######
df <- Envelope_data 

plot_per_well <- function(well_ids) {
  
  df_well <- df %>% 
    dplyr::filter(well == well_ids) # Filter for the current well

  g <- ggplot(df_well, aes(x = my_month, y = GL)) +
    geom_ribbon(aes(ymin = minGL, ymax = maxGL, fill = "Above/Below\nnormal range"), alpha = 0.5) +
    geom_ribbon(aes(ymin = Q20, ymax = Q80, fill = "Normal range\n(20-80th percentile)"), alpha = 1) +
    scale_fill_manual(values = c("dodgerblue4", "lightskyblue")) +
    geom_line(aes(x = my_month, y = GL, lty = "Groundwater\nLevel"), color = "black", size = 0.5) +
    scale_linetype_manual(values = c("dashed")) +
    labs(x = "Year", 
         y = "Groundwater levels (m)", 
         fill = " ") +
    theme(legend.position = "bottom",
          legend.direction = "horizontal",
          legend.title = element_blank()) +
    facet_wrap(~well, scales = "free") 
    #ggtitle(well_ids)  # Uncomment if you want a title per plot

  # Set the target directory for plots
  plots_dir <- file.path(rprojroot::find_rstudio_root_file(), 
                         "3_State_analysis", 
                         "Envelope_plots2")

  # Check if the directory exists; if not, create it
  if (!dir.exists(plots_dir)) {
    dir.create(plots_dir, recursive = TRUE)  
  }
  
  # Save the plot
  cowplot::save_plot(file.path(plots_dir, paste0(well_ids, ".png")), g)

  # Print progress message
  print(paste0("Generating plot for well ", well_ids))
  
  return(g)  # Return the plot (optional)
}

# Get unique well IDs
well_ids <- unique(df$well)
no_well <- length(well_ids)

# Loop through wells and generate plots
for(www in 1:no_well) {  
  plot_per_well(well_ids[www])
}

###############################################################################

```

## Seasonal groundwater conditions by aquifer

-   In this option to seasonal groundwater level conditions between 2019-2024 are summarised and percentage bar plots. This graph hasn't been used in the 2024 SoE reporting cycle but provide an alternative option

```{r , include=FALSE, cache=FALSE}
#################################################################

# Group monthly conditions by season and graph for each watyr year for each aquifer
Summary_plot <- ggplot(data_season_2019_2023_by_aquifer, aes(fill=con, y=per, x=season)) +
    geom_bar(position="fill", stat="identity")+
    scale_fill_manual(values = c("Above-normal"='cyan',
                               "Below-normal"='yellow',
                               "Highest-ever"='blue',
                               "Lowest-ever"='red',
                               "Normal"='green'))+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.15))+
  ylab("Percentage")+
  ggtitle("Monthly Groundwater level conditions grouped by Season and Year")+
  facet_grid(seayr~NAME,scales ="free_x")


#creates a Map folder
dir.create(file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Summary_plots"), recursive = TRUE, showWarnings = FALSE)

#Saves the pie chart plot to the created folder as a PNG file
ggsave(filename = file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Summary_plots", "Summary_plot.png"),
       plot = Summary_plot,
       width = 10, height = 8, dpi = 300)

```

## Facet maps

-   In this code chunk we create facet maps for the Heretaunga and Ruataniwha Plains based on seasonal groundwater level conditions between 2021-2024

```{r , include=FALSE, cache=FALSE}
#By season - note the seasons are different to the watyr by 1 month

map_sf_2021_2023_sea <- data_season_con %>%
  ungroup() %>%
  dplyr::filter(seayr %in% c(2021, 2022, 2023)) %>%
  na.omit() %>%
  mutate(con = fct_relevel(con, "Highest-ever", "Above-normal", "Normal", "Below-normal", "Lowest-ever")) %>%
  left_join(WellStor_sf_NZTM, by = "well") %>%
  st_as_sf(crs = 2193) %>%  # Ensure the result is an sf object with NZTM CRS
  st_intersection(allbasins) %>%
  dplyr::select(NAME, well, season, seayr, sea_gl, con) %>%
  dplyr::mutate(year_year = paste(seayr, sep = " / ")) %>%
  dplyr::filter(NAME %in% c("Heretaunga Plains", "Ruataniwha Plains"))
# 
#Map
tmap::tmap_mode("plot")

#Here we define the background map for Hawke's Bay
Map_2021_2023_sea_facet_Rua <- tmap::tm_shape(RP_plains_sf)+
  tmap::tm_fill()+
  tmap::tm_shape(map_sf_2021_2023_sea)+
    tmap::tm_dots(col = "con",
                  palette=c("Above-normal"='deepskyblue',
                            "Below-normal"='yellow',
                            "Highest-ever"='blue',
                            "Lowest-ever"='red',
                            "Normal"='chartreuse4'),
                  #title=unique(allData_sf2$month),
                  title= " ",
                  stretch.palette = FALSE,
                  size = 1,
                  shape =21,
                  clustering = FALSE,
                  legend.show = T,
                  legend.is.portrait = F)+
    tmap::tm_facets(c("year_year","season"),free.coords = FALSE)+
    tmap::tm_layout(legend.outside = T,
                    legend.outside.position = "bottom",
                    legend.position = c(0.1,0.4),
                    legend.outside.size = 0.15,
                    legend.text.size = 2)

 
#creates a Map folder
dir.create(file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Maps"), recursive = TRUE, showWarnings = FALSE)

#saves file
tmap::tmap_save(Map_2021_2023_sea_facet_Rua, 
  file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Maps", "Map_2021_2023_sea_facet_Rua.png")
)

# HERETAUNGA PLAINS

Map_2021_2023_sea_facet_Here <- tmap::tm_shape(HP_plains_sf)+
    tmap::tm_fill()+
    tmap::tm_shape(map_sf_2021_2023_sea)+
    tmap::tm_dots(col = "con",
                  palette=c("Above-normal"='deepskyblue',
                            "Below-normal"='yellow',
                            "Highest-ever"='blue',
                            "Lowest-ever"='red',
                            "Normal"='chartreuse4'),
                  #title=unique(allData_sf2$month),
                  title= " ",
                  stretch.palette = FALSE,
                  size = 1,
                  shape =21,
                  clustering = FALSE,
                  legend.show = T,
                  legend.is.portrait = F)+
    tmap::tm_facets(c("year_year","season"),free.coords = FALSE)+
    tmap::tm_layout(legend.outside = T,
                    legend.outside.position = "bottom",
                    legend.position = c(0.1,0.4),
                    legend.outside.size = 0.15,
                    legend.text.size = 2,
                    outer.margins = 0.01)
# 
tmap::tmap_save(Map_2021_2023_sea_facet_Here, 
  file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Maps", "Map_2021_2023_sea_facet_Here.png")
)


```

## Pie plots - by seasons

```{r, include=FALSE, cache=FALSE}

# 1. Create a seasonal column from your datetime column and store in GL_per_1
GL_per_1 <- GL_per %>%
  mutate(month = month(datetime),
         season = case_when(
           month %in% c(12, 1, 2) ~ "Summer",
           month %in% c(3, 4, 5) ~ "Autumn",
           month %in% c(6, 7, 8) ~ "Winter",
           month %in% c(9, 10, 11) ~ "Spring"
         ),
         # Change the factor order to Winter, Spring, Summer, Autumn
         season = factor(season, levels = c("Winter", "Spring", "Summer", "Autumn"))
  )

# 2. Filter for Heretaunga and calculate average seasonal GL; store in heretaunga_seasonal_1 
heretaunga_seasonal_1 <- GL_per_1 %>%
  filter(NAME == "Heretaunga Plains") %>%
  group_by(well, season, hydro_year) %>%
  summarise(avg_GL = mean(GL, na.rm = TRUE), .groups = 'drop')

# 3. Calculate percentiles for each season using historical seasonal averages; store in heretaunga_seasonal_2
heretaunga_seasonal_2 <- heretaunga_seasonal_1 %>%
  group_by(well, season) %>%
  mutate(percentile = percent_rank(avg_GL) * 100) %>%
  ungroup() %>%
  mutate(category = case_when(
    percentile == 0 ~ "Lowest-ever",
    percentile <= 25 ~ "Below-normal",
    percentile <= 75 ~ "Normal",
    percentile < 100 ~ "Above-normal",
    percentile == 100 ~ "Highest-ever",
    TRUE ~ "Unknown"
  ))

# 4. Summarise the counts of each category for plotting; store in plot_data_seasonal_1
plot_data_seasonal_1 <- heretaunga_seasonal_2 %>%
  group_by(hydro_year, season, category) %>%
  summarise(count = n(), .groups = 'drop') %>%
  complete(hydro_year, season, category = c("Highest-ever", "Above-normal", "Normal", "Below-normal", "Lowest-ever"),
           fill = list(count = 0)) %>%
  mutate(category = factor(category, levels = c("Highest-ever", "Above-normal", "Normal", "Below-normal", "Lowest-ever"))) %>%
  group_by(hydro_year, season) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup() %>%
  filter(hydro_year %in% c(2021, 2022, 2023))

# 5. Generate pie charts by season and hydrological year with hydrological years on the left (rows) and seasons on the top (columns)
Pie_by_seasons_HP <- ggplot(plot_data_seasonal_1, aes(x = "", fill = category, y = count)) +
  geom_bar(position = "fill", stat = "identity", width = 1, colour = "white") +
  geom_text(aes(label = ifelse(percentage == 0, "", paste0(round(percentage * 100, 1), "%"))),
            position = position_fill(vjust = 0.5),
            size = 3.5,
            check_overlap = FALSE) +
  coord_polar(theta = "y") +
  facet_grid(hydro_year ~ season) +
  theme_void() +
  scale_fill_manual(name = NULL, values = c(
    "Highest-ever" = "dodgerblue",
    "Above-normal" = "skyblue",
    "Normal" = "forestgreen",
    "Below-normal" = "yellow",
    "Lowest-ever" = "red"
  )) +
  theme(strip.text.x = element_text(size = 14),
        strip.text.y = element_text(size = 14))

#creates a Map folder
dir.create(file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Pie_plots"), recursive = TRUE, showWarnings = FALSE)

#Saves the pie chart plot to the created folder as a PNG file
ggsave(filename = file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Pie_plots", "Pie_by_seasons_HP.png"),
       plot = Pie_by_seasons_HP,
       width = 10, height = 8, dpi = 300)

```

## Pie plots - by aquifer

Pie charts visualize the **proportion of wells** in each groundwater condition category (`Normal`, `Above-normal`, etc.), grouped by **season and aquifer**.

**Key Dataframes:**

-   **`plot_data_seasonal_1`**: Aggregates groundwater conditions by **season**.

-   **`plot_data_aquifer`**: Summarizes groundwater levels for different **aquifers**.

```{r, include=FALSE, cache=FALSE}
############################################
# For Regional Chapter (AQ analysis) ########
############################################

# Convert hydro_year to numeric and filter recent data from the GL_per dataset
aq_data <- GL_per %>% 
  mutate(hydro_year = as.numeric(hydro_year))

aq_latest_year <- 2023
aq_recent_years <- (aq_latest_year - 2):aq_latest_year
aq_recent <- aq_data %>% filter(hydro_year %in% aq_recent_years)

# Calculate percentiles and categorise for AQ data.
# Each GL value is compared to the full historical GL values for the same well in aq_data.
aq_percentile_summary <- aq_recent %>%
  group_by(well) %>%
  mutate(percentile_rank = sapply(GL, function(x) mean(aq_data$GL[aq_data$well == well] <= x) * 100),
         category = case_when(
           percentile_rank == 0 ~ "Lowest-ever",
           percentile_rank <= 25 ~ "Below-normal",
           percentile_rank <= 75 ~ "Normal",
           percentile_rank < 100 ~ "Above-normal",
           percentile_rank == 100 ~ "Highest-ever",
           TRUE ~ "Unknown"
         )) %>% 
  ungroup()

# Ensure all categories are included when summarising
aq_all_categories <- data.frame(category = c("Lowest-ever", "Below-normal", "Normal", "Above-normal", "Highest-ever"))

# Prepare data for plotting pie charts by location and water year
aq_plot_data <- aq_percentile_summary %>%
  group_by(NAME, hydro_year, category) %>%
  summarise(count = n(), .groups = 'drop') %>%
  complete(NAME, hydro_year, category = aq_all_categories$category, fill = list(count = 0)) %>% 
  mutate(category = factor(category, levels = c("Highest-ever", "Above-normal", "Normal", "Below-normal", "Lowest-ever")))

# Generate pie charts for each aquifer (excluding "Minor aquifers")
pie_by_aquifer <- aq_plot_data %>%
  filter(NAME != "Minor aquifers") %>%
  group_by(NAME, hydro_year) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup() %>%
  ggplot(aes(x = "", fill = category, y = count)) +
  geom_bar(position = "fill", stat = "identity", width = 1, colour = "white") +
  geom_text(aes(label = ifelse(percentage == 0, "", paste0(round(percentage * 100, 1), "%"))),
            position = position_fill(vjust = 0.5),
            size = 3.5,
            check_overlap = FALSE) +
  coord_polar(theta = "y") +
  facet_grid(NAME ~ hydro_year, switch = "y",
             labeller = labeller(hydro_year = c(
               "2021" = "2021-2022",
               "2022" = "2022-2023",
               "2023" = "2023-2024"
             ))) +
  theme_void() +
  scale_fill_manual(name = NULL, values = c(
    "Highest-ever" = "dodgerblue",
    "Above-normal" = "skyblue",
    "Normal" = "forestgreen",
    "Below-normal" = "yellow",
    "Lowest-ever" = "red"
  )) +
  theme(strip.text.x = element_text(size = 14))


# Create the output folder (if it doesn't already exist)
dir.create(file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Pie_plots"), 
           recursive = TRUE, showWarnings = FALSE)

# Save the pie chart plot to a PNG file
ggsave(filename = file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Pie_plots", "Pie_by_Aquifer.png"),
       plot = pie_by_aquifer,
       width = 10, height = 8, dpi = 300)

```

# Tables

## Table for Jess

-   In this code chunk we create a table summarising groundwater level conditions

```{r, include=FALSE, cache=FALSE}

############################################
# For Catchment Chapters (EMA analysis) ####
############################################

# Filter EMA data to dates before 2024-07-01
ema_data <- GL_per_1_EMA %>% 
  filter(datetime < "2024-07-01")

# Calculate recent water years (last 3 years) based on the 'watyr' column
ema_latest_year <- max(ema_data$watyr)
ema_recent_years <- (ema_latest_year - 2):ema_latest_year
ema_recent <- ema_data %>% filter(watyr %in% ema_recent_years)

# Calculate percentiles and categorise for EMA data.
# Here each GL value is compared to all historical GL values for the same well in ema_data.
ema_percentile_summary <- ema_recent %>%
  group_by(well) %>%
  mutate(percentile_rank = sapply(GL, function(x) mean(ema_data$GL[ema_data$well == well] <= x) * 100),
         category = case_when(
           percentile_rank == 0 ~ "Lowest-ever",
           percentile_rank <= 25 ~ "Below-normal",
           percentile_rank <= 75 ~ "Normal",
           percentile_rank < 100 ~ "Above-normal",
           percentile_rank == 100 ~ "Highest-ever",
           TRUE ~ "Unknown"
         )) %>% 
  ungroup()

# Ensure all categories are included when summarising
ema_all_categories <- data.frame(category = c("Lowest-ever", "Below-normal", "Normal", "Above-normal", "Highest-ever"))

# Create a summary table for water years 2021, 2022, 2023, grouped by EMA and water year
ema_percentile_summary_table <- ema_percentile_summary %>%
  filter(watyr %in% c(2021, 2022, 2023)) %>%
  group_by(EMA, watyr) %>%
  summarise(
    average_percentile = round(mean(percentile_rank, na.rm = TRUE), 2),
    median_percentile = round(median(percentile_rank, na.rm = TRUE), 2),
    lowest_ever_count = sum(category == "Lowest-ever"),
    lowest_ever_percentage = round((sum(category == "Lowest-ever") / n()) * 100, 2),
    below_normal_count = sum(category == "Below-normal"),
    below_normal_percentage = round((sum(category == "Below-normal") / n()) * 100, 2),
    normal_count = sum(category == "Normal"),
    normal_percentage = round((sum(category == "Normal") / n()) * 100, 2),
    above_normal_count = sum(category == "Above-normal"),
    above_normal_percentage = round((sum(category == "Above-normal") / n()) * 100, 2),
    highest_ever_count = sum(category == "Highest-ever"),
    highest_ever_percentage = round((sum(category == "Highest-ever") / n()) * 100, 2),
    total_count = n(),
    wells_count = n_distinct(well),
    .groups = 'drop'
  )

#creates a Map folder
dir.create(file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Tables"), recursive = TRUE, showWarnings = FALSE)

# Export the EMA summary table to CSV in the newly created folder
write.table(ema_percentile_summary_table, 
            file = file.path(rprojroot::find_rstudio_root_file(), "3_State_analysis", "Tables", "groundwater_percentiles_summary.csv"), 
            sep = ",", row.names = FALSE, quote = FALSE)

# No_unique_wells_tests <- MK_results_combined_1 %>% 
#   group_by(period,area) %>% 
#   distinct(well) %>% 
#   summarise(n=dplyr::n()) %>% 
#   pivot_wider(names_from = period, values_from = n)


```
